Historical aerial surveys map long-term changes of forest cover and structure in the central Congo Basin

### 2020-01-24 manuscript/main.Rmd
revision

abstract: |
   Land Use and Land Cover change (LULCC) of African rainforest contribute to global carbon emissions. Yet, most historical estimates of LULCC and their carbon emissions rely on non-spatially explicity data in the pre-satellite era (< 1972). Past studies have expanded this pre-satellite time frame with declassified satellite surveillance data or aerial surveys, but none cover the Congo Basin. Here, we use historical aerial survey photos to map the extent and structure of LULCC around Yangambi, central Congo Basin, in 1958. Our study leveraged Structure-from-Motion to generate a large orthomosaic covering 82800 ha, counting 733 million pixels, at a sub meter resolution and geo-referenced to ~4.7 $\pm$ 4.3 m. Primary forest in the orthomosaic was classified with a Convolutional Neural Network Deep Learning approach. We used these data to quantify LULCC, landscape and canopy complexity and scale above ground biomass between historical and contemporary condition. We show a shift from historical highly structured industrial deforestation in 1958 (16200 ha), to contemporary smallholder farming clearing (8800 ha), increasing landscape fragmentation and forest edges. Efforts to quantify canopy texture and their link to carbon storage had limited to no success. Our analysis provides an insight in key spatial and temporal patterns of deforestation and reforestation at a multi-decadal scale and provide a historical context to land-cover and land-use change spatial patterns for past and ongoing field research in the area.
keywords: |
  Aerial survey, data recovery, CNN, Deep Learning, SfM, Congo Basin
acknowledgement: |
  This research was supported through the Belgian Science Policy office COBECORE project (BELSPO; grant BR/175/A3/COBECORE). KH acknowledges funding from the European Union Marie Skłodowska-Curie Action (project number 797668). I'm grateful for the support of Thales D'Haulleville volunteering his time in scanning the images.
authorcontributions: |
  K.H. conceived and designed the study, analyzed the data, prepared figures and tables, authored, reviewed and approved the final draft of the manuscript. T.d.H. scanned all image data. E.K. and T.d.H. provided plot based AGC estimates. T.D., K.J., E.K, H.B., P.S., F.V.,S.M., M.A., J.V.D.B., H.V and L.W. reviewed the final manuscript.
conflictsofinterest: |
  The authors declare no conflict of interest. The founding sponsors had no role in the design of the study; in the collection, analyses, or interpretation of data; in the writing of the manuscript, an in the decision to publish the results.
output: rticles::mdpi_article
---

@@

Tropical ecosystem services are severely impacted by deforestation and forest degradation [@hansen2013; @houghton2012; @tyukavina2015]. Not only does tropical forest Land-Use and Land-Cover Change (LULCC) constitute 10 to 15% of the total global emissions [@vanderwerf2009], changes in forest fragmentation affect the forest structure and function. Strong fragmentation effects decreases the number of large trees along forest edges [@laurance2000; @magnago2015], while species composition and biodiversity are equally negatively affected [@poorter2006; @barlow2016; @vandeperre2018]. Estimates show that 31% of carbon emissions are caused by edge effects alone [@brinck2017].

As such, anthropogenic factors through LULCC heavily affect forest structure and functioning [@fauset2017; @brinck2017; @didham1999]. Accurate estimates of LULCC and forest canopy structure are therefore imperative to estimate carbon emissions and other ecosystem services [@houghton2012; @hansen2013]. Remote sensing products have been key input as they provide accurate spatial information to help estimate carbon emissions [@mitchard2018; @hansen2013]. High resolution aerial images provide scientists tools to monitor forest extent, structure and carbon emissions as canopy texture is linked to above ground biomass [@Ploton2012; @Couteron2005; @Barbier2010]. Yet, these estimates are often limited in time to recent decades [@defries2002; @houghton2012; @hansen2013; @achard2014].

Historical estimates **of Land-Use and Land-Cover (LULC)**, in the pre-satellite era (< 1972), exist but generally rely on non-spatially explicit data (i.e socio-economic data) [@houghton2012; @ramankutty1999; @defries2002; @sader1988]. Efforts have been made to use other geo-spatial data sources such as historical maps [@willcock2016], declassified CORONA satellite surveillance data across the US and central Brazil [@song2015] as well as aerial surveys in post world war II German [@nita2018]. Survey data across the African continent is less common, inaccessible or both. Some studies do exist, as @buitenwerf2012, @hudak1998 and @okeke2006 used aerial survey images to map vegetation changes in South African savannas and the Adulam Nature Reserve, respectively, whilst @frankl2015 and @nyssen2016 mapped the Ethiopian highlands of 1930's.

Despite the existence of large historical archives of aerial survey imagery for the central Congo Basin (Figure 1, Appendix Figure 3) as of yet no studies have valorized these data. The lack of a consistent valorization effort is unfortunate as the African rainforest is the second largest on Earth and covers ~630 million ha, representing up to 66 Pg of carbon storage [@Lewis2009b], and currently loses forest at an increasing pace [@butsic2015]. Given the impact of LULCC on the structure and functioning of central African tropical forests, and their influence on both carbon dynamics [@Kearsley2013] and biodiversity [@vandeperre2018], accurate long term reporting of historical forest cover warrants more attention [@willcock2016].

Here, we use a combination of historical aerial photography and contemporary remote sensing data to map long-term changes in the extent and structure of the tropical forest surrounding Yangambi, in the central Congo Basin, effectively linking the start of the antrophocene [@lewis2015] with current assessments. Yangambi was, and remains, a focal center of forest and agricultural research and development in the central Congo Basin. Past research in the region allows for thorough assessment of land-use and land-cover change from a multi-disciplinary point of view, confronting us with complex deforestation and land-use patterns.

@@

Data for the central Congo Basin region, surrounding Kisangani, were collected in several flights during the dry season of 1958 and 1959 (from 8/01/1958 to 20/02/1958 and from 28/12/1958 to 9/01/1959 respectively, see Appendix Figure 1) to generate topographic maps of the area, supervised by the "Institut Géographique du Congo Belge" in Kinshasa (then Léopoldville). Black-and-White infrared images (0.4 - 0.9 $\mu$m) were gathered along flight paths running mostly from west to east, between **09:00 and 11:00** local time. Along a flight path continuous images were taken using a Wild Heerburgg RC5a (currently Leica Geosystems) with an Aviogon lens assembly (114.83mm / f 5.6, with a 90\degree view angle) resulting in square photo negative of 180 by 180 mm. Flights were flown at an average absolute altitude of ~5200 m above sea level, covering roughly 18 530 km^2^ at an approximate scale of 1/40 000. The use of the integrated autograph system ensured timely acquisition of pictures with a precise overlap (~1/3) between images. This large overlap between images together with flight parameters would allow post-processing, using stereographs, to create accurate topographic maps. Original data from this campaign are stored in the [Royal Museum for Central Africa](https://www.africamuseum.be/en) in Tervuren, Belgium (Figure 1).

```{r echo=FALSE, out.width='75%', fig.align="center", fig.cap="A box of historical aerial photographs (left) and a single aerial photograph (right) showing part of the Congo river. Note the meta-data provided in the margins of the image such as acquisition time and fleight height and the solar glare on the Congo river."}
knitr::include_graphics('./figures/print_examples.jpg')
```

@@

Data was processed into a georeferenced orthomosaic using a Structure from motion (SfM, @ullman1979) approach implemented in [Agisoft Metashape](http://www.agisoft.com) version 1.5.2 (Agisoft LLC, St. Petersburg, Russia). An orthomosaic corrects remote sensing data to represent a perfectly downward looking image, free from perspective distortions due to topography and camera tilt. Using the SfM technique features, areas in images with a large degree of similarity, are matched across various images to reconstruct a three dimensional scene (topography) from two-dimensional image sequences. During the SfM analysis we masked clouds, glare or large water bodies such as the Congo river.

@@

The CNN model was trained for 100 epochs with a batch size of 30 using Adam optimization [@kingma2017], maximizing the Intersect-over-Union (IoU) using Sørensen–Dice and categorical cross-entropy loss functions. Data augmentation included random cropping to 320 pixel squares, random orientation, scaling, perspective, contrast and brightness shifts and image blurring. The optimized model was used to classify the complete orthomosaic using a moving window approach with a step size of 110 pixels and a majority vote **(> 50% agreement)** across overlapping areas to limit segmentation edge effects. **In addition we provide, raw pixel level classification agreement data for quality control purposes (see Data availability below).** We refer to Figure 4 for a synoptic overview of the full deep learning learning workflow.

@@

To map long term land-use and land-cover change in the Yangambi region we used the contemporary Global Forest Change version 1.6 (GFC, tile 10N-020E) map data [@hansen2013]. Using the **GFC** data we calculated the latest state of the forest with respect to the conditions at the start of 1958, 60 years earlier. In our analysis we only included **GFC** pixels which recorded no **forest** loss throughout the whole 2000 - 2018 period. As the resolution of the historical forest classification exceeds that of the GFC map we downsampled our historical forest cover data to 30 m GFC resolution **using a nearest neighbour approach**. We masked out all water bodies using the Global Forest Change survey data mask layer, and limited the analysis to the right bank of the Congo river. We provide summary statistics of historical and contemporary forest loss and regrowth.

deforestation (persistent forest loss after 1958), reforestation (regrowth after loss in 1958).

@@

We estimated above ground carbon (AGC) losses and gains over time using plot based averages of recent inventory data at permanent sampling plots in the area (Figure 2). We refer to @Kearsley2013 for the survey method and allometric relations used to scale the survey data. Unlike standard square 1 ha plots edge plots were set back 200m from forest edges and were 50x200 m, with the 50 m side of the plot along the forest edge and continuing 200 m into the forest (Appendix Table 1). We further confirmed that forest edge plots did not show a significantly different AGC compared to those of non-edged / mixed forest plots (Mann Whitney U test, p < 0.05). Thus it was not necessary to explicitly quantify changes in AGC caused by edge effects. Moreover, we used the mean values of the mixed forest as representative for potential AGC losses. Despite the challenges inherent in quantifying AGC for forest edges we mapped the total extent of the edges in the contemporary landscape. To align our landscape analysis with exploratory analysis of the survey data we used a buffer of 200 m to estimate the extent of forest edges and patches, up to the location of forest edge plots.

Surveys of old plantations show a large variation in AGC, depending on age and the crop type. For example, the AGC values varied from 168.67 to 86.55 g Mg C ha^-1^, for *Hevea brasiliensis* (rubber tree) and *Elaeis guineensis* (oil palm) plots respectively (@bustillo2018, personal communications). These higher values are in line with the mixed AGC estimates in the area, while the palm plantations resemble old-regrowth values (81.8 Mg C ha^-1^, see **Appendix Table 1**). We therefore use both the estimates of old-regrowth and mixed forest to estimate AGC for regrowth. We did not have sufficient data to account for individual changes in AGC across plantations.

@@

We compared the structure of the canopy both visually and using Fourier Transform Textural Ordination (FOTO, @couteron2002). Fourier Transform Textural Ordination uses a principal component analysis (PCA) on radially averaged 2D Fourier spectra to characterize canopy (image) texture. The FOTO technique was first described by @couteron2002 to quantify canopy stucture in relation to biomass and biodiversity, and can be used across multiple scenes using normalization [@Barbier2010].

@@

For both site based and scene analysis we correlate PC values with permanent sample plot inventory data such as stem density, above ground biomass and tree species richness. Additional comparisons are made between contemporary Geo-Eye data and the historical orthomosaic derived PC values. Due to the few available permanent sampling plots in both scenes we used a non-parametric paired signed rank (Wilcoxon) test to determine differences between the PC values of the Geo-Eye and historical orthomosaic image scenes across mono-dominant and mixed forest types. In all analysis, mono-dominant site 4 was removed from the analysis due to cloud contamination.

@@

Our analysis provides a first spatially explicity historical composite of aerial survey images **in support of mapping land-use and land-cover** within the Congo Basin. The use of high resolution historical images combined with structure from motion image processing techniques allowed us to mosaic old imagery across a large extent. The final orthomosaic composition of the Yangambi region **provided** an image scene covering approximately 733 million pixels across **~82800 ha** (~23x36 km, Figure 2). The overall accuracy of the SfM orthomosaic composition was 0.88 m/pixel using the sparse cloud DEM for corrections at 45.8 m/pixel. The resulting georeferenced scene had a spatial accuracy of approximately **23 m**. Further georeferencing outside the SfM workflow reduced the mean error at the ground control points to 5.3 $\pm$ 4.9 px (~4.7 $\pm$ 4.3 m), with a median error of 2.9 px (2.6 m). The orthomosaic served as input for all subsequent LULCC analysis with all derived maps provided with the manuscript repository (see data & code availability statements below).

@@

**The CNN deep learning classifier reached an Intersection-over-Union (IoU) accuracy of 97% on the detection of disturbed forest in the out-of-sample (validation) synthetic landscape data. Using all pixels within the validatoin polygons (Figure 5) showed a similar accuracy value of ~98%. Using all polygons across the scene, including those used in the generation of testing and training synthetic landscapes, increased the accuracy to ~99% (Table 2). A comparison with recent pan-chromatic Geo-Eye data shows good agreement, with an accuracy of ~87% across all pixels, between the landsat based GFC data and downscaled CNN results (Table 2 and Figure 7).**

@@

## Long term LULC changes and Above Ground Carbon

@@

Recent deforestation follows a distincly different pattern compared to historical patterns. Historical deforestation showed a classical fishbone pattern for forest clearing with very sharp edges, while current patterns are patchy and ad-hoc (**figure 6C**, dark blue and green colours respectively). These differences are reflected in the analysis of landscape metrics of forest loss. Between the historical and contemporary LULCC maps we see an increase in small disturbances, as indicated by the decreasing area of the mean patch size, down to ~1.86 $\pm$ 0.75 ha from ~5.25 $\pm$ 5.02 ha historically. Perimeter lengths were longer historically, at 1451 $\pm$ 943 m,  compared to contemporary landscapes ~921 $\pm$ 362 m (Table 3). This shift in perimeter area ratio led to a similar change in the fractal index, increasing in value to 1.1 $\pm$ 0.05 from 1.09 $\pm$ 0.04 over time. Values closer to a fractal index of 2 suggest a more complex (fragmented) landscape.

@@

Changes in both land-use and land-cover led to concomitant changes in above ground carbon stocks. Recovery throughout the region was characterized for patches of forest and plantations. Assuming high density stands, based on previous work, this could amount to a carbon gains of 1592 Gg C across our study area, offsetting more recent losses of approximately 1408 Gg C. On the other hand, at the low end, if we assume a lower carbon density of 81.8 Mg C ha^-1^ this would result in a total carbon gain of 811 Gg C. Using our approach results indicate that overall deforestation around Yangambi has resulted in a loss of ~2416 Gg C in AGC stocks.

```{r echo = FALSE, eval = TRUE}
change_stats %>%
  kable(caption="Land use land cover change statistics of forest cover around Yangambi in the central Congo Basin. The data evaluates a difference between a historical (1958) aerial photography based survey and the Hansen et al. 2013 based satellite remote sensing data. Spatial coverage statistics are provided in square kilometers (km) and hectares (ha), rounded to the nearest integer as well as Above Ground Carbon (AGC) scaled using recent survey measurements.",
       booktabs = TRUE,
       digits = 3,
       row.names = TRUE,
       escape = FALSE,
       col.names = c("ha",
                     "(Gg C $ha\\textsuperscript{-1}$")
         ) %>%
    add_header_above(c(" " = 2,"AGC" = 1)) %>%
    kable_styling(latex_options = c("hold_position"))
```

@@

Visual interpretation of the scenes provide evidence that most locations do not change dramatically with respect to canopy composition, except for the large areas of disturbances in contemporary fallow or young-regrowth plots. One marked difference is noted in the mono-dominant site 6 (Appendix Table 1). Here, the current mono-dominant *Brachystegia laurenttii* is a recent development, changing the canopy structure visibly during the last half century (**Figure 8**). The previous varied canopy structure gave way to a more dense and uniform canopy. This is reflected in a change of the FOTO PC value from 0.19 historically to its current value of 0.54. This historical value is similar to the mean of contemporary mono-dominant stands with PC averaging 0.34 $\pm$ 0.1, and is only slightly higher than historical values for a mixed forest (0.18 $\pm$ 0.08). The reverse pattern is seen in the contemporary PC values. Here, the value of 0.54 exceeds those of most mono-dominant stands (0.35 $\pm$ 0.08), and is even further removed from the values noted for mixed forests (0.12 $\pm$ 0.03).

@@

Using only small subsets around existing permanent sampling plots we show distinct differences between forest types, with PC values in both historical and contemporary imagery markedly higher for the mono-dominant forest types compared to all others (**Appendix Figure 4**). Provided that the young-regrowth and fallow permanent sampling plots have seen recent disturbance the Wilcoxon signed rank test on the mixed and mono-dominant plots between the historical and contemporary PC values did not show a significant difference (p > 0.05). Similarly, no significant different using PC values extracted from the whole scene analysis was noted (p > 0.05). Any relationships between contemporary Geo-Eye data and permanent sampling plot measurements of Above Ground Biomass, stem density and species richness were non-significant (p > 0.05, Appendix **Figures 5-7**).

@@

### Land-use, Land-Cover change classification & validation

@@

**Our approach uses broadly defined homogeneous polygons to construct a balanced dataset of synthethic landscapes. The methodology is analoguous to the use of sparse labelling as used by @buscombe2018 and contrasts with the standard methodologies which generally extract pixel (windows) [@song2015] or delineate land cover classes [@nita2018] to drive a classifier or analysis. More so, the use of heavy image augmentation during model training sidesteps texture representation issues which affect classification of image scenes with inconsistent illumination or sharpness [@hudak1998] or ad-hoc feature engineering [@song2015]. The use of synthetic landscapes allowed us to account for most, but not all, of the variability within our orthomosaic. Our analysis has shown that despite being trained on historical data our model could map contemporary forest cover in remote sensing data with similar spatial and spectral characteristics (Figure 7), suggesting that the classifier consistently across works both space and time. We acknowledge that the use of synthetic landscapes is limited by the available homogeneous areas to sample from and the number of classes. Yet, the latter should not be a constrained as previous research efforts have focussed on simple forest loss maps [@hansen2013].**

@@

## Land-use, Land-Cover change & Above Ground Carbon

>>>>

abstract: |
   Given the impact of tropical forest disturbances on atmospheric carbon emissions, biodiversity and ecosystem productivity accurate long term reporting of Land-Use and Land-Cover (LULC) change in the pre-satellite era (< 1972) is an imperative. Here, we use a combination of historical (1958) aerial photography and contemporary remote sensing data to map long-term changes in the extent and structure of the tropical forest surrounding Yangambi (DR Congo), in the central Congo Basin. Our study leveraged Structure-from-Motion and a novel Convolutional Neural Network based LULC classifier, using synthetic landscapes based image augmentation, to map historical forest cover across a large sub meter resolution orthomosaic (~82800 ha) geo-referenced to ~4.7 $\pm$ 4.3 m. A comparison with contemporary LULC data showed a shift from previously highly regular industrial deforestation of large areas, to discrete smallholder farming clearing, increasing landscape fragmentation but also providing opportunties for substantial forest regrowth. Efforts to quantify long term canopy texture changes and a link to above ground carbon had limited to no success. Our analysis provides methods and insights into key spatial and temporal patterns of deforestation and reforestation at a multi-decadal scale, providing a historical context for past and ongoing forest research in the area.
keywords: |
  Aerial survey, data recovery, CNN, Deep Learning, SfM, Congo Basin
acknowledgement: |
  This research was supported through the Belgian Science Policy office COBECORE project (BELSPO; grant BR/175/A3/COBECORE). KH acknowledges funding from the European Union Marie Skłodowska-Curie Action (project number 797668). I'm grateful for the support of Thales de Haulleville volunteering his time in scanning the images.
authorcontributions: |
  K.H. conceived and designed the study, analyzed the data, prepared figures, tables and authored the final draft of the manuscript. T.d.H. scanned all image data. E.K. and T.d.H. provided plot based AGC estimates. T.D., K.J., E.K, H.B., P.S., F.V.,S.M., M.A., J.V.D.B., H.V and L.W. reviewed the final manuscript.
conflictsofinterest: |
  The authors declare no conflict of interest. The founding sponsors had no role in the design of the study; in the collection, analyses, or interpretation of data; in the writing of the manuscript, an in the decision to publish the results.
output: rticles::mdpi_article
---

@@

Tropical ecosystem services are severely impacted by deforestation and forest degradation [@hansen2013; @houghton2012; @tyukavina2015]. Not only does tropical forest Land-Use and Land-Cover Change (LULCC) constitute 10 to 15% of the total global carbon emissions [@vanderwerf2009], changes in forest fragmentation affect the forest structure and function [@fauset2017; @brinck2017; @didham1999]. Strong fragmentation effects decreases the number of large trees along forest edges [@laurance2000; @magnago2015], while species composition and biodiversity are equally negatively affected [@poorter2006; @barlow2016; @vandeperre2018]. Estimates show that 31% of carbon emissions are caused by edge effects alone [@brinck2017].

Accurate estimates of LULCC and forest canopy structure are therefore imperative to estimate carbon emissions and other ecosystem services [@houghton2012; @hansen2013]. Remote sensing products have been key inputs in LULCC assessments as they provide accurate spatial information to help estimate carbon emissions [@mitchard2018; @hansen2013]. More so, high resolution aerial images provide scientists tools to monitor forest extent, structure and carbon emissions as canopy texture is linked to above ground biomass [@Ploton2012; @Couteron2005; @Barbier2010]. Yet, most of these estimates are limited in time to recent decades [@defries2002; @houghton2012; @hansen2013; @achard2014].

Historical estimates **of Land-Use and Land-Cover (LULC)**, in the pre-satellite era (< 1972), exist but generally rely on non-spatially explicit data (i.e socio-economic data) [@houghton2012; @ramankutty1999; @defries2002; @sader1988]. Efforts have been made to use other geo-spatial data sources such as historical maps [@willcock2016], declassified CORONA satellite surveillance data across the US and central Brazil [@song2015] as well as aerial surveys in post World War II German [@nita2018]. Survey data across the African continent is less common, inaccessible or both. Some studies do exist, as @buitenwerf2012 and @hudak1998 used aerial survey images to map vegetation changes in South African savannas, whilst @frankl2015 and @nyssen2016 mapped the Ethiopian highlands of 1930's.

Across the central Congo Basin most of these historical images were collected within the context of national cartographic efforts by the "Institut Géographique du Congo Belge" in Kinshasa (then Léopoldville), DR Congo. Despite the existence of large archives of aerial survey imagery (Figure 1, Appendix Figure 3), as of yet, no studies have valorized these data. The lack of a consistent valorization effort is unfortunate as the African rainforest is the second largest on Earth and covers ~630 million ha, representing up to 66 Pg of carbon storage [@Lewis2009b], and currently loses forest at an increasing pace [@butsic2015]. Given the impact of LULCC on the structure and functioning of central African tropical forests, and their influence on both carbon dynamics [@Kearsley2013] and biodiversity [@vandeperre2018], accurate long term reporting of historical forest cover warrants more attention [@willcock2016].

Here, we use a combination of historical aerial photography and contemporary remote sensing data to map long-term changes in the extent and structure of the tropical forest surrounding Yangambi (DR Congo), in the central Congo Basin, effectively linking the start of the antrophocene [@lewis2015] with current assessments. Yangambi was, and remains, a focal center of forest and agricultural research and development in the central Congo Basin. Past research in the region allows for thorough assessment of land-use and land-cover change from a multi-disciplinary point of view, confronting us with complex deforestation and land-use patterns.

@@

Data for the central Congo Basin region, surrounding Kisangani, were collected in several flights during the dry season of 1958 and 1959 (from 8/01/1958 to 20/02/1958 and from 28/12/1958 to 9/01/1959 respectively, see Appendix Figure 1) to generate topographic maps of the area, supervised by the "Institut Géographique du Congo Belge" in Kinshasa, DR Congo (then Léopoldville). Black-and-White infrared images (0.4 - 0.9 $\mu$m) were gathered along flight paths running mostly from west to east, between **09:00 and 11:00** local time. Along a flight path continuous images were taken using a Wild Heerburgg RC5a (currently Leica Geosystems) with an Aviogon lens assembly (114.83mm / f 5.6, with a 90\degree view angle) resulting in square photo negatives of 180 by 180 mm. Flights were flown at an average absolute altitude of ~5200 m above sea level, covering roughly 18 530 km^2^ at an approximate scale of 1/40 000. The use of the integrated autograph system ensured timely acquisition of pictures with a precise overlap (~1/3) between images. This large overlap between images together with flight parameters would allow post-processing, using stereographs, to create accurate topographic maps. Original data from this campaign are stored in the [Royal Museum for Central Africa](https://www.africamuseum.be/en) in Tervuren, Belgium (Figure 1).

```{r echo=FALSE, out.width='75%', fig.align="center", fig.cap="A box of historical aerial photographs (left) and a single aerial photograph (right) showing part of the Congo river. Note the meta-data provided in the right side margin of the image, such as acquisition time and fleight height."}
knitr::include_graphics('./figures/print_examples.jpg')
```

@@

Data was processed into a georeferenced orthomosaic using a Structure-from-motion (SfM, @ullman1979) approach implemented in [Agisoft Metashape](http://www.agisoft.com) version 1.5.2 (Agisoft LLC, St. Petersburg, Russia). An orthomosaic corrects remote sensing data to represent a perfectly downward looking image, free from perspective distortions due to topography and camera tilt. Using the SfM technique features, areas in images with a large degree of similarity, are matched across various images to reconstruct a three dimensional scene (topography) from two-dimensional image sequences. During the SfM analysis we masked clouds, glare or large water bodies such as the Congo river.

@@

The CNN model was trained for 100 epochs with a batch size of 30 using Adam optimization [@kingma2017], maximizing the Intersect-over-Union (IoU) using Sørensen–Dice [@dice1945] and categorical cross-entropy loss functions. Data augmentation included random cropping to 320 pixel squares, random orientation, scaling, perspective, contrast and brightness shifts and image blurring. The optimized model was used to classify the complete orthomosaic using a moving window approach with a step size of 110 pixels and a majority vote **(> 50% agreement)** across overlapping areas to limit segmentation edge effects. **In addition, we provide raw pixel level classification agreement data for quality control purposes (see Data availability below).** We refer to Figure 4 for a synoptic overview of the full deep learning workflow.

@@

To map long term LULCC in the Yangambi region we used the contemporary Global Forest Change version 1.6 (GFC, tile 10N-020E) map data [@hansen2013]. Using the **GFC** data we calculated the latest state of the forest with respect to the conditions at the start of 1958, 60 years earlier. In our analysis we only included **GFC** pixels which recorded no **forest** loss throughout the whole 2000 - 2018 period. As the resolution of the historical forest classification exceeds that of the GFC map we downsampled our historical forest cover data to 30 m GFC resolution **using a nearest neighbour approach**. We masked out all water bodies using the Global Forest Change survey data mask layer, and limited the analysis to the right bank of the Congo river. We provide summary statistics of historical and contemporary forest loss and regrowth. **We map persistent forest loss after 1958, regrowth after loss in 1958, recent forest loss and long term (stable) forest cover.**

@@

We estimated above ground carbon (AGC) losses and gains over time using plot based averages of recent inventory data at permanent sampling plots in the area (Figure 2). We refer to @Kearsley2013 for the survey method and allometric relations used to scale the survey data. Unlike standard square 1 ha plots, edge plots were set back 200 m from forest edges and were 50x200 m, with the 50 m side of the plot along the forest edge and continuing 200 m into the forest (Appendix Table 1). We further confirmed that forest edge plots did not show a significantly different AGC compared to those of non-edged / mixed forest plots (Mann Whitney U test, p < 0.05). Thus it was not necessary to explicitly quantify changes in AGC caused by edge effects. Moreover, we used the mean values of the mixed forest as representative for potential AGC losses. Despite the challenges inherent in quantifying AGC for forest edges we mapped the total extent of the edges in the contemporary landscape. To align our landscape analysis with exploratory analysis of the survey data we used a buffer of 200 m to estimate the extent of forest edges and patches, up to the location of forest edge plots.

Surveys of old plantations show a large variation in AGC, depending on age and the crop type. For example, the AGC values varied from 168.67 to 86.55 Mg C ha^-1^, for *Hevea brasiliensis* (rubber tree) and *Elaeis guineensis* (oil palm) plots respectively (@bustillo2018, personal communications). These higher values are in line with the mixed AGC estimates in the area, while the palm plantations resemble old-regrowth values (81.8 Mg C ha^-1^, see **Appendix Table 1**). We therefore use both the estimates of old-regrowth and mixed forest to estimate AGC for regrowth. We did not have sufficient data to account for individual changes in AGC across plantations.

@@

We compared the structure of the canopy both visually and using Fourier Transform Textural Ordination (FOTO, @couteron2002). FOTO uses a principal component analysis (PCA) on radially averaged 2D Fourier spectra to characterize canopy (image) texture. The FOTO technique was first described by @couteron2002 to quantify canopy stucture in relation to biomass and biodiversity, and can be used across multiple scenes using normalization [@Barbier2010].

@@

For both site based and scene analysis we correlated PC values with permanent sample plot inventory data such as stem density, above ground biomass and tree species richness. Additional comparisons are made between contemporary Geo-Eye data and the historical orthomosaic derived PC values. Due to the few available permanent sampling plots in both scenes we used a non-parametric paired signed rank (Wilcoxon) test [@wilcoxon1945] to determine differences between the PC values of the Geo-Eye and historical orthomosaic image scenes across mono-dominant and mixed forest types. In all analysis, mono-dominant site 4 was removed from the analysis due to cloud contamination.

@@

Our analysis provides a first spatially explicity historical composite of aerial survey images **in support of mapping land-use and land-cover** within the Congo Basin. The use of high resolution historical images combined with SfM image processing techniques allowed us to mosaic old imagery across a large extent. The final orthomosaic composition of the Yangambi region **provided** an image scene covering approximately 733 million pixels across **~82800 ha** (~23x36 km, Figure 2). The overall accuracy of the SfM orthomosaic composition was 0.88 m/pixel using the sparse cloud DEM for corrections at 45.8 m/pixel. The resulting georeferenced scene had a spatial accuracy of approximately **23 m**. Further georeferencing outside the SfM workflow reduced the mean error at the ground control points to 5.3 $\pm$ 4.9 px (~4.7 $\pm$ 4.3 m), with a median error of 2.9 px (2.6 m). The orthomosaic served as input for all subsequent LULCC analysis with all derived maps provided with the manuscript repository (see data & code availability statements below).

@@

**The CNN deep learning classifier reached an Intersection-over-Union accuracy of 97% on the detection of disturbed forest in the out-of-sample (validation) synthetic landscape data. Using all pixels within the validatoin polygons (Figure 5) showed a similar accuracy value of ~98%. Using all polygons across the scene, including those used in the generation of testing and training synthetic landscapes, increased the accuracy to ~99% (Table 2). A comparison with recent pan-chromatic Geo-Eye data shows good agreement, with an accuracy of ~87% across all pixels, between the landsat based GFC data and downscaled CNN results (Table 2 and Figure 7).**

@@

## Long term changes in LULC and Above Ground Carbon

@@

Recent deforestation follows a distinctly different pattern compared to historical patterns. Historical deforestation showed a classical fishbone pattern for forest clearing with very sharp edges, while current patterns are patchy and ad-hoc (**figure 6C**, dark blue and green colours respectively). These differences are reflected in the analysis of landscape metrics of forest loss. Between the historical and contemporary LULCC maps we see an increase in small disturbances, as indicated by the decreasing area of the mean patch size, down to ~1.86 $\pm$ 0.75 ha from ~5.25 $\pm$ 5.02 ha historically. Perimeter lengths were longer historically, at 1451 $\pm$ 943 m,  compared to contemporary landscapes ~921 $\pm$ 362 m (Table 3). This shift in perimeter area ratio led to a similar change in the fractal index, slightly increasing in value to 1.1 $\pm$ 0.05 from 1.09 $\pm$ 0.04 over time. Values closer to a fractal index of 2 suggest a more complex (fragmented) landscape.

@@

Changes in both land-use and land-cover led to concomitant changes in AGC stocks. Recovery throughout the region was characterized for patches of forest and plantations. Assuming high density stands, based on previous work, this could amount to a carbon gains of 1592 Gg C across our study area, offsetting more recent losses of approximately 1408 Gg C. On the other hand, at the low end, if we assume a lower carbon density of 81.8 Mg C ha^-1^ this would result in a total carbon gain of 811 Gg C. Using our approach results indicate that overall deforestation around Yangambi has resulted in a loss of ~2416 Gg C in AGC stocks.

```{r echo = FALSE, eval = TRUE}
change_stats %>%
  kable(caption="Land use land cover change statistics of forest cover around Yangambi in the central Congo Basin. The data evaluates a difference between a historical (1958) aerial photography based survey and the Hansen et al. 2013 based satellite remote sensing data. Spatial coverage statistics are provided in square kilometers (km) and hectares (ha), rounded to the nearest integer as well as Above Ground Carbon (AGC) scaled using recent survey measurements.",
       booktabs = TRUE,
       digits = 3,
       row.names = TRUE,
       escape = FALSE,
       col.names = c("ha",
                     "Gg C $ha\\textsuperscript{-1}$")
         ) %>%
    add_header_above(c(" " = 2,"AGC" = 1)) %>%
    kable_styling(latex_options = c("hold_position"))
```

@@

Visual interpretation of the scenes provide evidence that most locations do not change dramatically with respect to canopy composition, except for the large areas of disturbances in contemporary fallow or young-regrowth plots. One marked difference is noted in the mono-dominant plot 6 (Appendix Table 1). Here, the current mono-dominant *Brachystegia laurentii* is a recent development, changing the canopy structure visibly during the last half century (**Figure 8**). The previous varied canopy structure gave way to a more dense and uniform canopy. This is reflected in a change of the FOTO PC value from 0.19 historically to its current value of 0.54. This historical value is similar to the mean of contemporary mono-dominant stands of *Gilbertiodendron dewevrei* with PC averaging 0.34 $\pm$ 0.1, and is only slightly higher than historical values for a mixed forest (0.18 $\pm$ 0.08). The reverse pattern is seen in the contemporary PC values. Here, the value of 0.54 exceeds those of most mono-dominant stands (0.35 $\pm$ 0.08), and is even further removed from the values noted for mixed forests (0.12 $\pm$ 0.03).

@@

Using only small subsets around existing permanent sampling plots we show distinct differences between forest types, with PC values in both historical and contemporary imagery markedly higher for the mono-dominant forest types compared to all others (**Appendix Figure 4**). Provided that the young-regrowth and fallow permanent sampling plots have seen recent disturbance the Wilcoxon signed rank test on the mixed and mono-dominant plots between the historical and contemporary PC values did not show a significant difference (p > 0.05). Similarly, no significant different using PC values extracted from the whole scene analysis was noted (p > 0.05). Any relationships between contemporary Geo-Eye data and permanent sampling plot measurements of Above Ground Carbon, stem density and species richness were non-significant (p > 0.05, Appendix **Figures 5-7**).

@@

### LULC classification & validation

@@

**Our approach uses broadly defined homogeneous polygons to construct a balanced dataset of synthethic landscapes. The methodology is analoguous to the use of sparse labelling as used by @buscombe2018 and contrasts with the standard methodologies which generally extract pixel (windows) [@song2015] or delineate land cover classes [@nita2018] to drive a classifier or analysis. More so, the use of heavy image augmentation during model training sidesteps texture representation issues which affect classification of image scenes with inconsistent illumination or sharpness [@hudak1998] or ad-hoc feature engineering [@song2015]. The use of synthetic landscapes allowed us to account for most, but not all, of the variability within our orthomosaic. Our analysis has shown that despite being trained on historical data our model could map contemporary forest cover in remote sensing data with similar spatial and spectral characteristics (Figure 7), suggesting that the classifier consistently works across both space and time. We acknowledge that the use of synthetic landscapes is limited by the available homogeneous areas to sample from and the number of classes. Yet, the latter should not be a constrained as previous research efforts have focussed on simple forest loss maps [@hansen2013].**

@@

### Long term changes in LULC and Above Ground Carbon
